---
title: "Gapminder Problem Set"
subtitle: "Data Exploration with SQL"
author: "YOUR NAME HERE"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


# Data preparation

## Create table in RSQLite

You first create a table called `gapminder` in your database. 

```{r}

# install.packages("RSQLite")
# install.packages("gapminder")

library(DBI)
library(gapminder)

# Connection to database SQlite
con <- dbConnect(RSQLite::SQLite(), ":memory:")

# Write data "gapminder" into database
dbWriteTable(con, "gapminder", gapminder)

# List tables
dbListTables(con)

```


# Data exploration 


## Selecting single columns

* Select the column `country`.

```{sql connection=con}


```


--- 
 
## Selecting multiple columns

* Select the columns `country`and population (`pop`).

```{sql connection=con}

```


---


## Select all columns

* Select all columns.

```{sql connection=con}

```


---

## Select all columns but not all observations

* Select all columns but only 5 observations.

```{sql connection=con}

```


---


## Select unique values

* a) Select all the unique values from the column `continent`. 

```{sql connection=con}

```


* b) Select all the unique values from the column `country`. 

```{sql connection=con}

```


---

## Count rows

* a) Count the number of rows in the dataset. 

```{sql connection=con}

```


* b) Count the number of distinct values in the column country.

```{sql connection=con}

```

---

## Filtering 

### Filtering of numeric values

* a) Obtain all countries with a population less than 1.000.000. 

```{sql connection=con}

```

* b) Obtain all distinct countries with a population less than 1.000.000. 

```{sql connection=con}

```

* c) Obtain all observations with a population less than 1.000.000. 

```{sql connection=con}

```

* d) Obtain all distinct observations with a population less than 1.000.000. 

```{sql connection=con}

```

* e) Count all observations with a population less than 1.000.000. 

```{sql connection=con}

```

* f) Obtain the columns `countries` and `continent`  with a population less than 1.000.000. 

```{sql connection=con}

```

* g) Obtain the distinct columns `countries` and `continent` with a population less than 1.000.000. 

```{sql connection=con}

```


---

### Filtering of text values

* a) Obtain all observations from Germany.

```{sql connection=con}

```


* b) Obtain distinct countries from Europe.

```{sql connection=con}

```


---

## WHERE: AND, OR, NOT

### AND 

* a) Obtain countries from continent Asia where population is less than 1.000.000

```{sql connection=con}

```


* b) Obtain distinct countries from continent Asia where population is less than 1.000.000

```{sql connection=con}

```


* c) Obtain distinct countries from continent Europe where the life expectancy is greater than 70 in the years before 1950.

```{sql connection=con}

```


* d) Obtain distinct countries from continent Europe where the life expectancy is greater than 70 in the years before 1960.

```{sql connection=con}

```

* e) Obtain distinct countries from continent Europe where the life expectancy is greater than 70 in the years before 1970.

```{sql connection=con}

```

---


### OR

* a) Obtain distinct countries and life expectancies where the life expectancy is less than 40 or greater than 75.

```{sql connection=con}

```


* b) Obtain distinct countries where the life expectancy in the year 1951 or 1952 is greater than 70. 

```{sql connection=con}

```


* c) Obtain distinct countries where the life expectancy is greater than 70 and the country is located in the continent Europe or Asia.

```{sql connection=con}

```

---



### BETWEEN

* a) Obtain distinct countries where life expectancy is greater or equal to 60 and less or equal to 70 (use operator `AND`).

```{sql connection=con}

```


* b) Obtain distinct countries where life expectancy is between 60 and 70 (use operator `BETWEEN`).

```{sql connection=con}

```



* c) Obtain distinct countries where life expectancy is between 60 and 70 and continent equals Africs. (use operator `BETWEEN`).

```{sql connection=con}

```


---

### WHERE IN

* a) Obtain distinct countries and years where year equals 1952, 1962, 1972, 1982 or 1992 (use operator `OR`).  

```{sql connection=con}

```

* b) Obtain distinct countries and years where year is in 1952, 1962, 1972, 1982 or 1992 (use operator `IN`).  

```{sql connection=con}

```

---


### NULL 

* a) Obtain the names of distinct countries which values are not missing in the gapminder table.

```{sql connection=con}

```

* b) Obtain the names of distinct countries which values of life expectancy are not missing in the gapminder table.

```{sql connection=con}

```

---


### LIKE and NOT LIKE

* a) Obtain the names of distinct countries where the continent starts with the uppercase letter "E" (use operator `%`). 

```{sql connection=con}

```


* b) Obtain the names of distinct countries where the continent starts the word "Europ" (use operator `LIKE` and `_`).

```{sql connection=con}

```


* c) Obtain the names of distinct countries where the continent does not start the letter "E" (use operator `NOT LIKE` and `%`).

```{sql connection=con}

```


---

# Aggregate Functions

## SUM, AVG, MAX, MIN

* a) Add up the values of the column GDP per capita (`gdppercap`) where continent is Asia. 

```{sql connection=con}

```

* b) Calculate the average of the variable life expectancy.

```{sql connection=con}

```

* c) Obtain the highest value of the column life expectancy.  

```{sql connection=con}

```

* d) Obtain the lowest value of the column life expectancy.  

```{sql connection=con}

```

* e) Obtain the average value of the column life expectancy for continent Asia.

```{sql connection=con}

```

* f) Obtain the average value of the column life expectancy where country begins with the uppercase letter "G".

```{sql connection=con}

```

* g) Obtain the highest value of the column life expectancy where country contains the word "Germa" (use the operator _)

```{sql connection=con}

```

* h) Obtain the average value of the column life expectancy between the years 1952 and 1962.

```{sql connection=con}

```

---


## Aliasing 

* a) Obtain the average value of the column life expectancy (call the resulting column "average_lifeexp") as well as the highest value of life expectancy (call the resulting column "maximum_lifeexp"). Use the operator `AS` to obtain the desired results.

```{sql connection=con}

```

* b) Obtain the sum of the column GDP per capita (call the resulting column "gdppercap_1952") for the year 1952. 

```{sql connection=con}

```

* c) Obtain the sum of the column GDP per capita (call the resulting column "gdppercap_1992") for the year 1992.

```{sql connection=con}

```

* d) Obtain the highest value for the colum year.  

```{sql connection=con}

```


* e) Obtain the lowest value for the colum year.

```{sql connection=}

```



# Sorting and grouping

## ORDER BY

* a) Obtain the data for the columns `country` and population (`pop`) in the year 2007 and order by population (ascending). 

```{sql connection=con}

```

* b) Obtain the data for the columns `country` and population (`pop`) in the year 2007 and order by population (descending). 

```{sql connection=con}

```

* c) Obtain the data for the columns `country` and `continent` and order by continent (ascending). 

```{sql connection=con}

```

* d) Obtain the data for the columns `country` and `continent` and order by continent (descending). 

```{sql connection=con}

```

* e) Obtain the data for the columns `country`,  `continent` and `pop` where continent is not starting with a capital "A" and order the result by population  (ascending). 

```{sql connection=con}

```

* f) Obtain the data for the columns `country`,  `continent` and `pop` where continent is not starting with a capital "A" and order the result by population  (descending). 

```{sql connection=con}

```

* g) Obtain the data for the columns `country`,  `continent` and `pop` where the year is in 2002 or 2007 (use the operator `IN`) and order the result by population (ascending). 

```{sql connection=con}

```



## Sorting multiple columns

* Obtain the data for the columns `country`,  `continent` where the year is 2007 and order the result by continent and country (ascending). 


```{sql connection=con}

```



## GROUP BY

* a) Obtain the data for the column `continent`, count all cases by continent and order the result (ascending). 
```{sql connection=con}

```

* b) Obtain the average life expectancy for the column `continent`. Group the results by `continent`.  

```{sql connection=con}

```

* c) Obtain the average life expectancy for the column `continent`. Group by continent and order the results by average life expectancy (ascending).  

```{sql connection=con}

```

* d) Obtain the average life expectancy for the column `continent` for the year 2007. Group by continent and order the results by average life expectancy (ascending).  

```{sql connection=con}

```


---

## Having

* a) Obtain the average life expectancy (AS avg_lifeexp) and average GDP per capita (AS avg_gdppercap) by continent  for the year 2007 where the total population is greater than 900000000 and order by continent.


```{sql connection=con}

```

* b) Obtain the average life expectancy (AS avg_lifeexp) and average GDP per capita (AS avg_gdppercap) by continent  for the year 2007 where the number of countries is greater than 50.


```{sql connection=con}

```



---

# Joins

- Import data from the world happiness report 2020 (csv "whr_20.csv") and call the table "happy":

```{r connection=con}

library(tidyverse)

happy_df <- read_csv("https://raw.githubusercontent.com/kirenz/datasets/master/whr_20.csv") 

happy <- 
  happy_df %>% 
  rename(
        country = country_name, 
        happiness = ladder_score,
        gdp = logged_gdp_per_capita,
        family = explained_by_social_support,
        liefeexpectancy = explained_by_healthy_life_expectancy,
        trust = explained_by_perceptions_of_corruption
  ) %>% 
  select(country, happiness, gdp, family, liefeexpectancy, trust)
    
dbWriteTable(con, "happy", happy)

dbListTables(con)

```


- Let's have a look at the data:

```{sql connection=con}
SELECT *
FROM HAPPY;
```


## Inner Join

* a) Inner join the `gapminder` table on the left to the `happy` table on the right, aliasing gapminder as "g" and happy as "h", select `country` and `gdppercap` from the gapminder table and `happiness` from the happy table for the year 2007:

```{sql connection=con}

```

* b) Repeat the query from a) with `USING()` instead of `ON`: 

```{sql connection=con}

```

* c) Now we would like to obtain a table with three columns: country, the population in 2002 and the population in 2007. 

Therefore, join (`SELF JOIN`) the table gapminder with itself on `country`. SQL does not allow same-named fields, so alias gapminder as g1 and g2. Select g1.pop as pop2002 and g2.pop as pop2007.

```{sql connection=con}

```

* d) Add a new column to our table in c): `growth` (difference between population in 2002 and 2007).

Use a SELECT statement to calculate a new column, aliased as growth, that calculates the population growth from 2002 to 2007 for each country, using g2.pop and g1.pop.

```{sql connection=con}

```

* e) Now calculate the percentage population growth from 2002 to 2007 for each country, and alias it as `growth_perc`. You might want to use the `CAST` function to convert data from integer to a numeric type. Use the `ROUND` function to round growth_perc to two decimal places. 

```{sql connection=con}

```


---

## Left Join

* a) Perform an `INNER JOIN` with the gapminder table on the left and the happy table on the right. Alias the gapminder table as "g" and the happy table as "h". Select `country`, `continent` and `lifeexp` from gapminder and `happiness` from happy. Join on country and filter for the year 2007 in the gapminder table. Order by `lifeexp` ascending. Note how many records are in the result of the join in the query result tab.

```{sql connection=con}

```

* b) Change the code of a) to perform a `LEFT JOIN` instead of an `INNER JOIN`. After executing this query, note how many records the query result contains. 

```{sql connection=con}

```

---


## Subqueries

* a) Use the happy table to discover which countries have higher than overall average happiness (use a subquery)? 

```{sql connection=con}

```

* b) Obtain the overall average happiness: 

```{sql connection=con}

```
