---
title: "Introduction to SQL"
subtitle: "Joins"
author: "Prof. Dr. Jan Kirenz"
institute: "HdM Stuttgart"
output:
  xaringan::moon_reader:
    css: ["footer-header.css", "default", "default-fonts"]
    lib_dir: libs
    nature:
      titleSlideClass: [center, middle]
      highlightLines: true
      highlightStyle: github
      countIncrementalSlides: false
---
layout: true
  
<div class="my-footer"><span> Introduction to SQL | Prof. Dr. Jan Kirenz </span></div> 

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(connection = "con")
```

???

---

# PostgreSQL Setup

```{r, include=FALSE}
pw = ""
```

```
pw = "your_password"
```

```{r}
library(DBI)
library(RPostgres)

con <- dbConnect(RPostgres::Postgres(),
                 dbname = "postgres",
                  host = "localhost", 
                  port = 5432,
                  user = "postgres", 
                  password = pw)
```

---
class: inverse, center, middle

### The examples in this presentation are based on the excellent book "A Beginner's Guide to Storytelling with Data" from Anthony DeBarros (2018). 

---
# JOIN example: Departments & Employees

## CREATE the departments table:

```{sql connection=con}
CREATE TABLE departments (
    dept_id bigserial,
    dept varchar(100),
    city varchar(100),
    CONSTRAINT dept_key PRIMARY KEY (dept_id),
    CONSTRAINT dept_city_unique UNIQUE (dept, city)
);
```


--

* `dept_id` column is the table's primary key

* Table includes a **UNIQUE** constraint

 * Guarantees that values in a column are unique. 

 * It requires that each row have a unique pair of values for *dept* and *city*
 
* We add these constraints to avoid duplicate data.
 
---
class: inverse, center, middle

## PRIMARY KEY

### A primary key is a column whose values uniquely identify each row in a table.


### This column has certain contstraints: it must have a unique value for each row and it can't have missing values.

---

## CREATE employees table

```{sql connection=con}
CREATE TABLE employees (
    emp_id bigserial,
    first_name varchar(100),
    last_name varchar(100),
    salary integer,
    dept_id integer REFERENCES departments (dept_id),
    CONSTRAINT emp_key PRIMARY KEY (emp_id),
    CONSTRAINT emp_dept_unique UNIQUE (emp_id, dept_id)
);
```

---

## CREATE employees table

- `emp_id` is the table's **primary key**: it uniquely identifies each row in the employees table

- `dept_id` (we added it as a constraint when creating the table) is called a **foreign key** 

 - It requires a value entered in a column to already exist in the primary key of the table it **references**.
 
 -  Values in `dept_id` in the employees table must exist in `dept_id` in the departments table

- UNIQUE: each row must have a **unique** pair of emp_id and dept_id

---

class: inverse, center, middle

## FOREIGN KEY 

### Unlike a primary key, a **foreign key** column can be empty or it can contain duplicate values

---
## INSERT values


```{sql connection=con}
INSERT INTO departments (dept, city)
VALUES
    ('Tax', 'Atlanta'),
    ('IT', 'Boston');
```

--

```{sql connection=con}
INSERT INTO employees (first_name, last_name, salary, dept_id)
VALUES
    ('Nancy', 'Jones', 62500, 1),
    ('Lee', 'Smith', 59300, 1),
    ('Soo', 'Nguyen', 83000, 2),
    ('Janet', 'King', 95000, 2);
```

---

## SELECT values

```{sql connection=con, output.var="de_all"}
SELECT *
FROM departments;
```

--

```{r, echo=FALSE}
DT::datatable(
  head(de_all, 4),
  fillContainer = FALSE, options = list(pageLength = 4)
)
```

---
## SELECT values


```{sql connection=con, output.var="em_all"}
SELECT *
FROM employees;
```

--

```{r, echo=FALSE}
DT::datatable(
  head(de_all, 4),
  fillContainer = FALSE, options = list(pageLength = 4)
)
```

---

## JOIN the tables

```{sql connection=con, output.var="em_de_j"}
SELECT *
FROM employees JOIN departments
ON employees.dept_id = departments.dept_id;
```

--

```{r, echo=FALSE}
DT::datatable(
  head(em_de_j, 4),
  fillContainer = FALSE, options = list(pageLength = 4)
)
```

---

# JOIN Types

* Creating two tables to explore JOIN types

```{sql connection=con}
CREATE TABLE schools_left (
    id integer CONSTRAINT left_id_key PRIMARY KEY,
    left_school varchar(30)
);
```

--

```{sql connection=con}
CREATE TABLE schools_right (
    id integer CONSTRAINT right_id_key PRIMARY KEY,
    right_school varchar(30)
);
```

---

### Insert values

```{sql connection=con}
INSERT INTO schools_left (id, left_school) VALUES
    (1, 'Oak Street School'),
    (2, 'Roosevelt High School'),
    (5, 'Washington Middle School'),
    (6, 'Jefferson High School');
```

--

```{sql connection=con}
INSERT INTO schools_right (id, right_school) VALUES
    (1, 'Oak Street School'),
    (2, 'Roosevelt High School'),
    (3, 'Morrison Elementary'),
    (4, 'Chase Magnet Academy'),
    (6, 'Jefferson High School');
```

---

# JOIN

- we use JOIN or INNER JOIN, when we want to return rows that have a match in the columns we used for the join

--

```{sql connection=con, output.var="sc_j"}
SELECT *
FROM schools_left JOIN schools_right
ON schools_left.id = schools_right.id;
```

--

```{r echo=FALSE}
knitr::kable(sc_j, format = 'html')
```

---

# INNER JOIN

```{sql connection=con, output.var="sc_ij"}
SELECT *
FROM schools_left INNER JOIN schools_right
ON schools_left.id = schools_right.id;
```

--

```{r echo=FALSE}
knitr::kable(sc_ij, format = 'html')
```

---

# JOIN or INNER JOIN

- Here the join includes all columns in both tables (*).

- Then we specify the two tables to join around the JOIN keyword.

- At last we specify which columns we're joining on, here the id columns of both tables.

- Three school IDs match in both tables, JOIN or INNER JOIN returns only the three rows of those IDs that match.

- Values (schools) that exist only in one of the two tables don't appear in the result.

- Notice that the columns from the left table display on the left of the result table.

- Use JOIN or INNER JOIN when you're working with well-structured, well-maintained data sets and only need to find rows that exist in all the tables you're joining.

---

# INNER JOIN with USING

```{sql connection=con, output.var="sc_ij_id"}
SELECT *
FROM schools_left INNER JOIN schools_right
USING (id);
```

--

```{r echo=FALSE}
knitr::kable(sc_ij_id, format = 'html')
```

---

# LEFT JOIN

- LEFT JOIN returns all rows from the left table and displays blank rows from the right table if no matching values are found in the joined columns.

--

```{sql connection=con, output.var="sc_left_j"}
SELECT *
FROM schools_left LEFT JOIN schools_right
ON schools_left.id = schools_right.id;
```

--

```{r echo=FALSE}
knitr::kable(sc_left_j, format = 'html')
```


---
class: inverse, center, middle

## RIGHT JOIN & LEFT JOIN

### Use either of these join types when you want your query results to contain **all** the rows from one of the tables... 

### ... or when you want to look for **missing values** in one of the tables.

---

# RIGHT JOIN

- RIGHT JOIN returns all rows from the right table and displays blank rows from the left table if no matching values are found in the joined columns

--

```{sql connection=con, output.var="sc_right_j"}
SELECT *
FROM schools_left RIGHT JOIN schools_right
ON schools_left.id = schools_right.id;
```

--

```{r echo=FALSE}
knitr::kable(sc_right_j, format = 'html')
```

---

# Aliasing (AS)

* Hint: Instead of writing the full table name, you can use table aliasing as a shortcut:

```{sql connection=con, output.var="al_1"}
SELECT e.salary, d.city
FROM employees AS e
INNER JOIN departments AS d
ON e.dept_id = d.dept_id;
```

--

```{r echo=FALSE}
knitr::kable(al_1, format = 'html')
```

---

# FULL OUTER JOIN

- Shows **all** rows from both tables in a join, regardless of whether any match:

```{sql connection=con, output.var="sc_outer1"}
SELECT *
FROM schools_left FULL OUTER JOIN schools_right
ON schools_left.id = schools_right.id;
```

--

```{r echo=FALSE}
knitr::kable(sc_outer1, format = 'html')
```


---

## FULL OUTER JOIN

```{sql connection=con, output.var="sc_outer2"}
SELECT *
FROM schools_right FULL OUTER JOIN schools_left
ON schools_right.id = schools_left.id;
```

--

```{r echo=FALSE}
knitr::kable(sc_outer2, format = 'html')
```


---

# FULL OUTER JOIN with USING


```{sql connection=con, output.var="sc_outer1_use"}
SELECT *
FROM schools_left FULL OUTER JOIN schools_right
USING (id);
```

--

```{r echo=FALSE}
knitr::kable(sc_outer1_use, format = 'html')
```

---

# CROSS JOIN


```{sql connection=con, output.var="sc_cross"}
SELECT *
FROM schools_left CROSS JOIN schools_right; 
```

* Lines up each row in the left table with each row in the right table to present all possible combinations of row. 

* Because the join doesn't need to find matches between key fields, there is no need to provide the clause using ON or USING.

---
class: center, middle

```{r echo=FALSE}
knitr::kable(sc_cross, format = 'html')
```


---

# Using NULL to find missing values

```{sql connection=con, output.var="sc_null"}
SELECT *
FROM schools_left LEFT JOIN schools_right
ON schools_left.id = schools_right.id
WHERE schools_right.id IS NULL;
```

--

```{r echo=FALSE}
knitr::kable(sc_null, format = 'html')
```

- The result shows only the one row from the left table that didn't have a match on the right side.


---

class: center, middle



```{sql connection=con, include=FALSE}
DROP TABLE employees;
```


```{sql connection=con, include=FALSE}
DROP TABLE schools_left;
```


```{sql connection=con, include=FALSE}
DROP TABLE schools_right;
```


```{sql connection=con, include=FALSE}
DROP TABLE departments;
```


```{r include = FALSE}
dbDisconnect(con)

```





## Thank you!

[Prof. Dr. Jan Kirenz](https://www.kirenz.com)  

HdM Stuttgart  
Nobelstraße 10  
70569 Stuttgart

![](https://upload.wikimedia.org/wikipedia/commons/5/5e/HdM_Logo.svg)
